---
title: "Movie_Ratings_Statistics_Step_2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory.
setwd("C:/Users/Juan Caballero/Desktop/dataSciencePortfolio/02movieRatingsStatistics")

# Load libraries needed.
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Importing Data.
Dataset obtained from: https://www.kaggle.com/datasets/stefanoleone992/rotten-tomatoes-movies-and-critic-reviews-dataset (Data has been scraped from the publicly available website https://www.rottentomatoes.com as of 2020-10-31.)

I have two file formats, .csv and .xlsx. Importing .xlsx (slower to read) to main original data types.

```{r}
# rotten_tomatoes_df <- read.csv("rotten_tomatoes_movies.csv")
rotten_tomatoes_df <- read_excel("rotten_tomatoes_movies.xlsx")

summary(rotten_tomatoes_df)
```

## Data Preparation and Cleaning.

``` {r}
# Check for missing values.
sum(is.na(rotten_tomatoes_df$tomatometer_rating))
sum(is.na(rotten_tomatoes_df$audience_rating))
sum(is.na(rotten_tomatoes_df$original_release_date))

# Drop rows with missing critic or audience ratings.
cleaned_rotten_tomatoes_data <- rotten_tomatoes_df %>% 
  drop_na(tomatometer_rating, audience_rating, original_release_date)

# Select relevant columns
columns_selected <- c('movie_title', 'content_rating', 'genres', 'directors', 'actors',
                      'original_release_date', 'runtime', 'production_company',
                      'tomatometer_rating',
                      'audience_rating')
cleaned_rotten_tomatoes_data <- cleaned_rotten_tomatoes_data %>% select(all_of(columns_selected))

# Extract year from release date
cleaned_rotten_tomatoes_data <- cleaned_rotten_tomatoes_data %>%
  mutate(release_year = as.numeric(format(original_release_date, "%Y")))

# Create score discrepancy variable
cleaned_rotten_tomatoes_data <- cleaned_rotten_tomatoes_data %>%
  mutate(score_discrepancy = tomatometer_rating - audience_rating)
```

## Display Cleaned Dataset.

``` {r}
# Display structure of cleaned dataset
glimpse(cleaned_rotten_tomatoes_data)

# Display summary of cleaned dataset
summary(cleaned_rotten_tomatoes_data)

# Aggregate the data by year to get mean critic and audience scores.
trend_data <- cleaned_rotten_tomatoes_data %>%
  group_by(release_year) %>%
  summarize(mean_critic_score = mean(tomatometer_rating, na.rm = TRUE),
            mean_audience_score = mean(audience_rating, na.rm = TRUE))

# Plot trends in critic vs. audience scores over time.
ggplot(trend_data, aes(x = release_year)) +
  geom_line(aes(y = mean_critic_score, color = "Critic Score"), linewidth = 1) +
  geom_line(aes(y = mean_audience_score, color = "Audience Score"), linewidth = 1) +
  labs(title = "Trends in Critic vs. Audience Scores Over Time",
       x = "Year of Release",
       y = "Mean Score",
       color = "Score Type") +
  scale_color_manual(values = c("Critic Score" = "blue", "Audience Score" = "red")) +
  theme_minimal()

# Plot trend lines
ggplot(trend_data, aes(x = release_year)) +
  geom_smooth(aes(y = mean_critic_score, color = "Critic Score"), method = "loess", se = FALSE) +
  geom_smooth(aes(y = mean_audience_score, color = "Audience Score"), 
              method = "loess", se = FALSE) +
  labs(title = "Trends in Critic vs. Audience Scores Over Time",
       x = "Year of Release",
       y = "Mean Score",
       color = "Score Type") +
  scale_color_manual(values = c("Critic Score" = "blue", "Audience Score" = "red")) +
  theme_minimal()

# Plot scatter plots.
ggplot(trend_data, aes(x = release_year)) +
  geom_point(aes(y = mean_critic_score, color = "Critic Score"), size = 2) +
  geom_point(aes(y = mean_audience_score, color = "Audience Score"), size = 2) + 
  labs(title = "Trends in Critic vs. Audience Scores Over Time", x = "Year of Release", 
       y = "Mean Score", color = "Score Type") + 
  scale_color_manual(values = c("Critic Score" = "blue", "Audience Score" = "red")) + theme_minimal()

# Plot scatter plots with trend lines
ggplot(trend_data, aes(x = release_year)) +
  geom_point(aes(y = mean_critic_score, color = "Critic Score"), size = 2) +
  geom_point(aes(y = mean_audience_score, color = "Audience Score"), size = 2) +
  geom_smooth(aes(y = mean_critic_score, color = "Critic Score"), method = "loess", se = FALSE) +
  geom_smooth(aes(y = mean_audience_score, color = "Audience Score"), method = "loess", se = FALSE) +
  labs(title = "Trends in Critic vs. Audience Scores Over Time",
       x = "Year of Release",
       y = "Mean Score",
       color = "Score Type") +
  scale_color_manual(values = c("Critic Score" = "blue", "Audience Score" = "red")) +
  theme_minimal()

```

## What information is not self-evident?
Means of critic and audience scores are grouped by movies' release year.

Rotten Tomatoes website was launched on August 1998.

Critic and audience scores rate movies dating back to 1914.

The specific impact of different genres (a movie can be classified in more that one genre), directors, and actors (more than one) on the score discrepancies.

The influence of budget and box office performance on these scores (No provided in dataset).

## What are different ways you could look at this data?
Analyzing score discrepancies by genre and release year.

Investigating specific directors or actors with consistent score gaps.

Comparing blockbuster movies to independent films.

## How do you plan to slice and dice the data?
Group data by genre, director, and release year.

## How could you summarize your data to answer key questions?
Use summary statistics (mean, median, standard deviation).
Create pivot tables for aggregating scores by genre, director, etc.

## What types of plots and tables will help you to illustrate the findings to your questions?
Bar plots for score discrepancies by genre.

Line plots to show trends over time.

Box plots to compare score distributions between blockbusters and indie films.

## Do you plan on incorporating any machine learning techniques to answer your research questions? Explain.
Right now, I am not sure which machine learning techniques I can use. Week 10 will cover machine learning concepts. 

## Questions for future steps.
How and where can I get movie budget, and box office performance.

How to filter genre, director, actor, 

How to arrange data by different categories to reveal patterns.

Reduce data to focus on key variables relevant to the research questions.

Create new variables to measure and compare differences.

Summary statistics highlighting trends and differences across categories.

Using pipe operator to streamline data manipulation and transformation steps.
